Evaluation of Cache-Oblivious Data Structures [draft-1]

### 2008-05-26 paper/paper.tex
More updates to the paper including some pictures of data structures.

% TODO: ensure all references in paper.bib are cited in the paper!
% TODO: format checklist (see e-mail)
\documentclass{acm_proc_article-sp}

@@

% TODO: more meaningful section titles?

@@

% Copy and shorten introduction from proposal
% -> Motivate research into cache-efficient algorithms
% -> Move "classification" subsection describing three models to front
% -> Describe each of the three models
% -> Introduce three data structures:
%     - note that hash tables have poor locality of reference
%     - note that btrees have good locality of reference but require pagesize
% -> Finally, present goal of the research (incorporate "Research Questions" in
% proposal)

% Marielle suggested:
%  - give the "standard" memory model an explicit name
%    (isn't "cache-unaware" enough?)
%  - move examples of cache oblivious data structures to front
%  - mention examples of cache unaware and cache aware algorithms
%    (at least Btree, tries and hash tables are mentioned)

% Note explicitly that cache-oblivious algorithms are relatively complex
% resulting in higher constant consts, so that the main question is wether
% the improved locality of reference compensates for the overhead incurred.

@@

% TODO: picture?

Our hash table implementation uses a fixed size index which must be specified
when creating the hash table, as this simplifies the implementation
considerably. The index is stored at the front of the file, after which values
are simply appended in the order in which they are added. Note that we do not
support removing elements from the hash table, which means we do not have to
deal with holes that would otherwise occur in the stored file. For our
experiments the FNV-1a hash function \cite{noll2004fnv} is used (modulo the
size of the index) to map values to slots.

@@

B-tree pages are ordered in a tree structure. Values in a page are stored
in order, and pointers are added that originate between a pair of adjacent
values, and point to a subtree containing only values that fall in between those
values. The resulting tree is typically very shallow, which is beneficial,
as the number of pages that need to be retrieved (worst case) is equal to
the height of the tree.

When a new value does not fit in the page where it is to be inserted,
the page is split into two pages; the median value from the old page is
moved to the parent page, and two new pages (containing the values less
than respectively greater than the median) are created. When the top-most
page needs to be split, a new (empty) top-level page is created and the height
of the B-tree is increased by one. As a result, all leaf nodes in a B-tree are
at the same depth, and all pages stored are at least half full.

@@

The implementation of the Bender set consists of two parts: a sparse array
storing the values in order, and a binary tree stored in van Emde Boas
layout (similar to the one proposed by Prokop) that is used to find values
efficiently.
Both the tree and the array may contain special \emph{empty values}.
A depiction of a Bender set storing four values (with a capacity for eight) is
given in Figure \ref{fig-bender-set}.

\begin{figure}
\centering
\includegraphics{bender-set}
\caption{A depiction of a Bender set}\label{fig-bender-set}
\end{figure}

A Bender set is created with a capacity $C$ that is a power of two.
Initially, the array stores only empty values and is partitioned into
\emph{windows} on several \emph{levels}.
On the highest level, there is a single window of size $C$. Each subsequent level
has twice as many windows of half that size, and the lowest level has size
$log_2 C$ (rounded up to the next power of 2).
For each level a maximum \emoph{density} is chosen, with the lowest level having
density 1, the highest level having density $\delta$, and the density for
the intermediary levels obtained by interpolating linearly between 1 and
$\delta$.

@@

In Table \ref{tab-cases} an overview is given of the parameters of the models used
and the resulting properties of the state space search.
Note that the number of transitions relative to the number of iterations indicates
the percentage of insertion operations that results in a new element being added
to the set, which ranges approximately from 4 to 2.7 non-updating queries per
insertion.

@@

\section{Results}

TODO: describe parameters of data structures
Explain results for B-tree (higher page size packs more data), hash table (larger index, less collisions, cache miss anyway)

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-queuesize}
\caption{Queue size}
\label{fig-quesize}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-transitions}
\caption{Transitions}
\label{fig-transitions}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-bender_wctime}
\caption{Bender set execution time on Eratosthenes}
\label{fig-eratosthenes-bender_wctime}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-bender_vss}
\caption{Bender set memory usage on Eratosthenes}
\label{fig-eratosthenes-bender_vss}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-hash_wctime}
\caption{Hash table execution time on Eratosthenes}
\label{fig-eratosthenes-hash_wctime}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-hash_vss}
\caption{Hash table memory usage on Eratosthenes}
\label{fig-eratosthenes-hash_vss}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-btree_wctime}
\caption{B-tree execution time on Eratosthenes}
\label{fig-eratosthenes-btree_wctime}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-eratosthenes-btree_vss}
\caption{B-tree memory usage on Eratosthenes}
\label{fig-eratosthenes-btree_vss}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-leader2_wctime}
\caption{Execution times on Leader2}
\label{fig-leader2_wctime}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-leader2_vss}
\caption{Memory usage on Leader2}
\label{leader2_vss}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-peterson_wctime}
\caption{Execution time on Peterson\_N}
\label{fig-peterson_wctime}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=83mm]{fig-peterson_vss}
\caption{Memory usage on Peterson\_N}
\label{fig-peterson_vss}
\end{figure}

% Present results!
% Tables, graphs, etc.
% What to include:
%   queue size over iterations
%       for each program
%
%   insertions per query:
%       for each program
%
%   per program:
%       wctime over iterations
%          for each set
%       memory over iterations
%          for each set

\section{Discussion}
% Explain results, draw conclusions.
% - relate to insert/test ratio
% - relate to hardware (not many levels of hierarchy)

@@

\bibliographystyle{plain}
\bibliography{paper}
\end{document}

>>>>

\documentclass{acm_proc_article-sp}

@@

@@

@@

\begin{figure}
\centering
\includegraphics[width=83mm]{hash-table}
\caption{A depiction of a separate-chaining hash table}
\label{fig-hash-table}
\end{figure}

Our hash table implementation uses a fixed size index which must be specified
when creating the hash table, as this simplifies the implementation
considerably. The index is stored at the front of the file, after which values
are simply appended in the order in which they are added. Note that we do not
support removing elements from the hash table, which means we do not have to
deal with holes that would otherwise occur in the stored file.
Figure \ref{hash-table} shows a hash table (with an index size of four,
storing three values) and the way it is stored in consecutive memory.

For our
experiments the FNV-1a hash function \cite{noll2004fnv} is used (modulo the
size of the index) to map values to slots.

@@

\begin{figure}
\centering
\includegraphics[width=83mm]{b-tree}
\caption{A depiction of a B-tree}
\label{fig-b-tree}
\end{figure}

B-tree pages are ordered in a tree structure. Figure \ref{fig-b-tree} depicts a
B-tree of height two storing eight values.
In each page the values are stored in lexicographical order, and the values
in the first leaf page are lexicographically smaller than ``John'', the values
in the second leaf page are between ``John'' and ``Philip'' and the values in
the third leaf page are greater than ``Philip''. Note that since the page
size is fixed, not all pages can be completely filled.

Since every page can store many values, the resulting tree is typically very
shallow, which is beneficial as the number of pages that need to be retrieved
is equal to the height of the tree (worst case).

To insert a value, simply find the leaf page where it belongs. If this page
does not have enough free space to insert it, the page is split into two pages;
the median value from the old page is moved to the parent page, and two new pages
(containing the values less than respectively greater than the median) are
created. When the top-most page needs to be split, a new (empty) top-level page is
created and the height of the B-tree is increased by one. As a result, all leaf
nodes in a B-tree are at the same depth, and all pages stored are at least half
full, except possibly the root page.

@@

The implementation of the Bender set consists of two parts: a sparse array
storing the values in order, and a binary tree stored in van Emde Boas
layout (similar to the one proposed by Prokop) that is used to find values
efficiently.
Both the tree and the array may contain special \emph{empty values}.
A depiction of a Bender set storing four values (with a capacity for eight) is
given in Figure \ref{fig-bender-set}, with the window population counts for
the sparse array given on the left (density thresholds not shown) and the index tree for the same array on the right.

\begin{figure}
\centering
\includegraphics[width=83mm]{bender-set}
\caption{A depiction of a Bender set}
\label{fig-bender-set}
\end{figure}

A Bender set is created with a capacity $C$ that is a power of two.
Initially, the array stores only empty values and is partitioned into
\emph{windows} on several \emph{levels}.
On the highest level, there is a single window of size $C$. Each subsequent level
has twice as many windows of half that size, and the lowest level has size
$log_2 C$ (rounded up to the next power of 2).
For each level a maximum \emph{density} is chosen, with the lowest level having
density 1, the highest level having density $\delta$, and the density for
the intermediary levels obtained by interpolating linearly between 1 and
$\delta$.

@@

% TODO: add state sizes to table?

In Table \ref{tab-cases} an overview is given of the parameters of the models used
and the resulting properties of the state space search.
Note that the number of transitions relative to the number of iterations gives an
indication of the ratio of value look-ups (insertions of values that are already
present in the set) and actual insertions, ranging from approximately from 4 to
2.7 look-ups per insertion.

@@

\section{Results}
Each of the tree data structures has some configurable parameters. The hash table
needs to be parametrized with the size of the index, the B-tree with the size of
the pages, and the Bender set with the density parameter ($\delta$). To determine
which parameters to use, various different parameters were first tried on the
first (and smallest) test case. In this case, a million relatively small states
need to be stored.

Figure \ref{fig-eratosthenes-hash_wctime} shows the execution times for the hash
table. About a million states need to be stored and as expected, the hash table
with an index size that is close to this number performs best. The hash table
with a small index (100,000 slots) performs well initially, but gets slower as
it becomes too full, at which point each slot in the hash table has to store a
long list of values that map to that slot.
The hash table with a large index (10,000,000 slots) is a bit slower than the hash
table with an index size of 1,000,000, possible because of cache misses caused by
the larger index size.
As Figure \ref{fig-eratosthenes-hash_vss} shows, the only difference in memory
usage by the different hash tables is in the constant size of the index.
Despite not being the fastest in this case, we will use the hash table with an
index size of 10,000,000 for further experiments, as other test cases require
storing well over 1 million states and we want to use a single size that works
well for all test cases.

Figure \ref{fig-eratosthenes-btree_wctime} and Figure
\ref{fig-eratosthenes-btree_vss} show the execution time and memory usage
for the B-tree with page sizes ranging from 1KB to 16KB. The execution platform
maps memory in 4KB pages, which would suggest that using pages less than 4KB
makes little sense. Indeed, the B-tree with 1KB pages seems to perform worst,
but if we take the memory use into account, this is more likely due to the fact
that fewer items fit on a single page (which means a relatively large portion of
the page remains unused) than anyting else.
The 4KB and 16KB pages are relatively close (both in execution time and memory
usage); we select the 16KB page size for further experiments because state sizes
are larger in other test cases, in which case the 4KB page size may cause similar
problems as the 1KB page size here.

Finally, in Figure \ref{fig-eratosthenes-bender_wctime} the execution times for
the Bender set with several different density parameters are given. With a lower
density, more space is required, but new insertions less frequently cause large windows to be rebalanced.
Suprisingly, the execution times for the different density values are comparable.
Figure \ref{fig-eratosthenes-bender_vss} shows that the lower density parameters do
indeed use a lot more memory. It appears that the advantage of low density is
negated by the overhead of constructing increasingly large data structures.
The set with $\delta=0.125$ cannot even finish the test case in the memory
that is available. Since execution times are similar, we select
$\delta=0.5$ for future experiments in order to minimize memory use.

% TODO: test with larger density!

Now that we have established the parameters to use for the other test cases.
The execution times for the three testcases are presented in figures
\ref{fig-eratosthenes_wctime}, \ref{fig-leader2_wctime} and
\ref{fig-peterson_wctime}, while the corresponding memory usage is presented in
figures \ref{fig-eratosthenes_vss}, \ref{fig-leader2_vss} and \ref{fig-peterson_vss} respectively.

\section{Discussion}
The three test cases paint a very similar picture: in all cases, the hash table
offers the best performance in terms of both execution time and memory usage,
followed by the B-tree. In all cases, the Bender set performs worst by a large
margin.

The difference in execution time between the hash table and B-tree increases as
the size of the states becomes larger. This is explained by the fact that the
height of the B-tree depends on the average number of values per page; larger
values means less values per page, and therefore a deeper tree, and as a
result more pages to be fetched for a query. The hash table does not have such a
limitation, as every value in the bucket must be fetched independently, regardless
of the size of these values.

When it comes to memory usage the B-tree and hash table have similar requirements;
the B-tree has slightly larger overhead per value stored, but initially the hash
table uses more memory because of the space allocated for the index.

The performance of the Bender set does not seem to depend on the page size,
and as a result the difference between the Bender set and the B-tree is smallest
when the page size is largest. Unfortunately, even then it requires about three
times as much time and a lot more memory.

The test cases used all have a relatively high percentage of insertions relative
to value look-ups. This may explain the good performance of the hash table (for
which insertions are barely more expensive than look-ups) as well as the bad
performance of the Bender tree (for which insertions are relatively expensive,
especially when a window is rebalanced). Unfortunately, the insert/look-up
ratio for our test cases varies too little to derive a meaningful conclusion
in this regard.

Finally, it should be noted that in our experiments only part of the memory
hierarchy was used (up to the use of main memory). This still involves several
levels of processor cache, but it is not a very deep hierarchy. With a deeper
memory hierarchy, with greater differences in access times between the levels,
the cache-efficient data structures (the B-tree and the Bender set) should
perform relatively better.

@@

\bibliographystyle{plain}
\bibliography{paper}

%\begin{figure}
%\centering
%\includegraphics[width=83mm]{fig-queuesize}
%\caption{Queue size}
%\label{fig-quesize}
%\end{figure}

%\begin{figure}
%\centering
%\includegraphics[width=83mm]{fig-transitions}
%\caption{Transitions}
%\label{fig-transitions}
%\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=83mm]{fig-eratosthenes-hash_wctime}
\caption{Execution time of hash table on Eratosthenes}
\label{fig-eratosthenes-hash_wctime}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=83mm]{fig-eratosthenes-btree_wctime}
\caption{Execution time of B-tree on Eratosthenes}
\label{fig-eratosthenes-btree_wctime}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=83mm]{fig-eratosthenes-bender_wctime}
\caption{Execution time of Bender set on Eratosthenes}
\label{fig-eratosthenes-bender_wctime}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=83mm]{fig-eratosthenes-hash_vss}
\caption{Memory usage of hash table on Eratosthenes}
\label{fig-eratosthenes-hash_vss}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=83mm]{fig-eratosthenes-btree_vss}
\caption{Memory usage of B-tree on Eratosthenes}
\label{fig-eratosthenes-btree_vss}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=83mm]{fig-eratosthenes-bender_vss}
\caption{Memory usage of Bender set on Eratosthenes}
\label{fig-eratosthenes-bender_vss}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=83mm]{fig-eratosthenes_wctime}
\caption{Execution time on Eratosthenes}
\label{fig-eratosthenes_wctime}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=83mm]{fig-leader2_wctime}
\caption{Execution time on Leader2}
\label{fig-leader2_wctime}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=83mm]{fig-peterson_wctime}
\caption{Execution time on Peterson\_N}
\label{fig-peterson_wctime}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=83mm]{fig-eratosthenes_vss}
\caption{Memory usage on Eratosthenes}
\label{fig-eratosthenes_vss}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=83mm]{fig-leader2_vss}
\caption{Memory usage on Leader2}
\label{fig-leader2_vss}
\end{figure}

\begin{figure}[p]
\centering
\includegraphics[width=83mm]{fig-peterson_vss}
\caption{Memory usage on Peterson\_N}
\label{fig-peterson_vss}
\end{figure}

% What else to include:
%   queue size over iterations
%       for each program
%
%   insertions per query:
%       for each program
%

\end{document}
